# Using MATLAB as a Powerful Calculator

MATLAB can do basic arithmetic, including trigonometry and linear algebra. Read through the MATLAB basics and try the examples below to get used to using MATLAB commands.


If you ever need additional help, recall the MathWorks help site: [here](http://www.mathworks.com/help/matlab/).

To get started, open MATLAB, create a new Live Script and save it as "Lab4.mlx".

## MATLAB Basics for calculation:

* We can add (`+`), subtract(`-`), multiply (`*`), and divide(`/`) any two numbers 
* All of our favorite functions are in MATLAB:
  + **The square root function**:
  
    i.e. To calculate $\sqrt{44944}$, just type `sqrt(44944)`
  + **The exponential function**:
  
    i.e. To calculate $e^4$, just type `exp(4)`
  + **The natural log function**: 
  
    i.e. To calculate $\ln{(3)}$, just type `log(3)`
    
    Note: $\log_{10}{(3)} = \cfrac{\ln{(3)}}{\ln{(10)}}$
  + **Trig functions**: 
  
    i.e. To calculate cos$\left(2\right)$ where $2$ is in radians, just type `cos(2)`.  To calculate $\cos{(2)}$ where $2$ is in degrees, type `cosd(2)`
  + **Powers**: 
  
    i.e. To calculate $8^{1.3}$, just type `8^(1.3)`

* All of our favorite numbers are there too:
    + For $\pi$, just type `pi`
    + For $\sqrt{-1}$, just type `i`
    
      Caution: If you assign the variable $i$ a value, it will overwrite this!
* Putting a semi-colon (`;`) at the end of any line in MATLAB will suppress the output - this means it will complete the operation, but it won't show you the output on your screen. 
  

```{block, type='rmdquestion'}
#### Question 1
  a) Find $\sqrt{3697929}$
  b) Note that MATLAB approximates when you are not dealing with a perfect square.  Approximate $\sqrt{134502}$
  c) Approximate $ln{(74)}$.
  d) Approximate $\log_{10}{8.6}$
  e) Let $x = 42$.  Approximate $sin{(x)}$ assuming $x$ is measured in radians.
  f) Let $x = 42$.  Approximate $sin{(x)}$ assuming $x$ is measured in degrees.
```

## Early words of caution:
* MATLAB is case sensitive!  If you define `b = 2` and then try to use `B`, MATLAB will tell you `B is not defined`.
* Anytime you define a variable (i.e. `x = 4`), that is stored for later use.  Once MATLAB is closed, all variables are erased.  To clear your variables at any time without closing MATLAB, you have two options:
  1) Clear an individual variable (i.e. `clear x`)
  2) Clear all variables (i.e. `clear all`)
* Every time you close MATLAB, your variables are cleared.  So, if you want to run commands that require variables you have already defined during a previous MATLAB session, you need to redefine them (i.e. rerun those lines of code) before you can run the new commands.
* When dealing with any matrix or vector of values, operations are interpreted as LINEAR ALGEBRA OPERATIONS (matrix operations)!  
  + We can only add/subtract two things that have the same number of elements
  + Multiplication is interpreted as matrix multiplication.  
  + If you tell MATLAB to perform an operation that does not make sense in the Linear Algebra setting, it will give you an error message.
  + We will see examples of this in the next section when dealing with vectors and matrices.
  

## Dealing with vectors/matrices:

Remember that a vector is simply a matrix with a single row or a single column – they are defined and interpreted in the same way.

### REVIEW: Defining a vector/matrix:
* Always use square brackets $\left[\dots\right]$ to define a vector/matrix
* Elements can be separated by a space or a comma
* Rows can be separated by a semi-colon or by hitting return without closing the vector/matrix
  + Examples:
    1) ```{r}
      A = [1  2  3; 4  5  6; 7  8  9]
      ```
    2) ```{r}       
      A = [1,2,3 
      4,5,6   
      7,8,9]
      ```
* To define a column vector, we can define it as a column vector or as the transpose (in MATLAB we use `'` to denote transpose) of the row vector.
  + Examples:
    1) ```{r}
      x = [1;2;3;4]
      ```
    2) ```{r}
      x = [1
      2
      3
      4]
      ```
    3) ```{r}
      x =[1  2  3  4]'
      ```
* To get a (row) vector consisting of a range of numbers, we can specify the endpoints, using a colon
  + Examples:
    1) If not given an increment, MATLAB will assume we want integers.
      ```{r}    
      v = 1:10
      ```
      creates the vector $v = \left[1,2,3,4,5,6,7,8,9,10\right]$
    2) We can tell MATLAB an increment to use.
      ```{r}
      v = 1:0.1:1.5
      ```
      creates the vector $v = \left[1,1.1,1.2,1.3,1.4,1.5\right]$

```{block, type='rmdquestion'}
#### Question 2
  - Create a vector $v$ containing all of the numbers from 0 to 1, in increments of 0.01.
  - Create another vector $w$ containing all of the numbers from 3 to 5, in increments of 0.02.
  - Create a third vector $u$ containing all of the integers from 1 to 100 (i.e. 1,2,3,...,100).
```

### Accessing Elements, Adding, and Subtracting
* To access an element of a vector/matrix, we use parentheses with the position of the element (starting at 1)
  + Examples:
    1) ```{r}
      x = [1;4;7;5];
      x(2)
      ```
      should give us the 2nd element, which is $4$.
      
    2) ```{r}
      A = [1  2  3; 4  5  6; 7  8  9]
      A(2,3)
      ```
      should give us the $(2,3)$ element, which is $6$.

* Operations are defined in the LINEAR ALGEBRA sense:
    + Adding/Subtracting two vectors/matrices:
      - They must have the same dimensions (number/arrangement of elements)
      - Adding/subtracting is done element-wise
      - Examples:
        1) ```{r}
        A = [1  2  3; 4  5  6; 7  8  9];
        B = [3  -2  5; -7  5  1; 0  -4  -6];
        A+B
        ```
        should give us the matrix 
$$\begin{bmatrix} 
4  & 0 & 8 \\ 
-3 & 10 & 7 \\
7 & 4 &  3 \\
\end{bmatrix}$$

        2) ```{r}
        A = [1  2  3; 4  5  6; 7  8  9];
        B = [3  -2  5; -7  5  1; 0  -4  -6];
        A-B
        ```
        should give us the matrix 
$$\begin{bmatrix} 
-2  & 4 & -2 \\ 
11 & 0 & 5 \\
7 & 12 &  15 \\
\end{bmatrix}$$

```{block, type='rmdquestion'}
#### Question 3
  - Using your vectors from Question 2, create a new vector, $z$, by adding $v$ and $w$.  What happens if you try to add $v$ and $u$?
  - Using MATLAB, determine the 22nd entry of $z$?
  - Given the following matrices,$A$ and $B$, compute $A+B$ and $B-A$.
  ```
  $$A = \begin{bmatrix} 
-1  & 6 & 0 \\ 
13 & 2 & -5 \\
9 & 1 &  -7 \\
\end{bmatrix}$$
  $$B = \begin{bmatrix} 
4 &  0 & 8 \\ 
-3 & 1 & 6 \\
7 & -4 &  3 \\
\end{bmatrix}$$

### Multiplying two vectors/matrices:
* This is done in the LINEAR ALGEBRA sense of MATRIX MULTIPLICATION ("middle dimensions must match")
* If we want to do multiplication element-wise, we must modify the command (using a period before the operation):

  `A*B` = matrix multiplication

  `A.*B` = element-wise multiplication

  This period trick works for any operation and is interpreted as "do this element-wise"

  If we want to square each term in a vector x, just type
  
  `x.^2`

  OR we can use a for loop



  ````markdown
  for i=1:length(x)
          x(i) = x(i)^2;
  end;
  ````
  
  We will learn more about using for loops later in this course!

```{block, type='rmdquestion'}
#### Question 4
  - Using the vector $z$ from Question 3, create a new vector, $y$, with each entry equal to the cube of the corresponding entry in $z$. 
  - Using MATLAB, determine the 22nd entry of $y$?
  - Using the matrices $A$ and $B$ from Question 3, create two new matrices, $C$ and $D$.  
    $C$ should be the matrix obtained using `matrix multiplication`.    
    $D$ should be the matrix obtained using `element-wise multiplication`.
  - Using MATLAB, determine the $(2,3)$ entry of $C$ and the $(2,3)$ entry of $D$.
  - Consider the matrix below, $F$.  Why is it impossible to multiple $A$ by $F$ element-wise?  
  - Again consider the matrix $F$.  Can you multiply $A$ times $F$ using matrix multiplication (in that order)?  What about $F$ times $A$?  
    You don't need Linear Algebra - use MATLAB to help answer this question.
```  
$$F = \begin{bmatrix} 
2  & -6  \\ 
3 & -5  \\
4 & -4  \\
\end{bmatrix}$$

## Running a simple MATLAB function
In your BIOL 101 lab this week, you measured the weight of either two types of maple seeds or maple leaves.  You then ran a Student's two-sample *t-test* on your data in Excel.  Now, lets use MATLAB to run that same statistical test and compare the output.

The ttest2() function in MATLAB is used to run a Student's two-sample *t-test*.  

Some helpful notes:

* The inputs to this function are two vectors, $x$ and $y$.
* If you run this function with a single output, it will return a test decision, $h$, for the null hypothesis.
The null hypothesis is that the data in vectors $x$ and $y$ comes from independent random samples with equal means and equal variances. The alternative hypothesis is that the data in $x$ and $y$ comes from populations with unequal means. \
$h=1$ if the test rejects the null hypothesis at the 5% significance level (i.e. $\alpha = 0.05%), and $h=0$ otherwise.
* If you run this function with two outputs, it will return the test decision, $h$, along with the p-value.
* For more information, you can always visit the [MathWorks website](http://www.mathworks.com/help/matlab/) and search for ttest2!\


```{block, type='rmdquestion'}
#### Question 5
 - Create four column vectors in MATLAB that contain your data from the two types of maple seeds/leaves.  Use variable names that will help you remember which vector contains which data (i.e. W1, W2, L1, L2).
 - Use the ttest2() function in MATLAB to run a Student's two-sample *t-test* on your data for each variable.  Make sure you are inputting the correct pair of vectors for each test.
 - Interpret the results in terms of the null hypotheses.
 - Do these results match those you obtained when you ran the Student's *t-test* in Excel?
```  

